---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# ch.fso.pasta

[![dataset](https://img.shields.io/endpoint?url=https://raw.githubusercontent.com/benjamin-arfa/ch.fso.pasta/main/data-raw/shield.json)](https://github.com/benjamin-arfa/ch.fso.pasta/actions/workflows/update_data.yaml)

The ch.fso.pasta package provides versioned time series data and their meta
information for scientific research. In addition, the package contains the
extract-transform-load (ETL) functionality that sources the data from its
original provider.

**Arrivals and overnight stays in holiday homes and collective accommodation, by
month** — the Swiss Federal Statistical Office's PASTA survey (statistics on
supplementary accommodation, the sibling of HESTA hotel accommodation), monthly
since 2017.

Unlike the other FSO archives in OpenTSI, this one reads **directly from
[stats.swiss](https://stats.swiss)** (Swiss Stats Explorer, SDMX REST) rather
than from the KOF Time Series Database, so it needs no API key.
`docs/PLAYBOOK.md` documents the recipe so it can be reused for any other
stats.swiss dataflow.

## Key structure

```
ch.fso.pasta.<accommodation>.<indicator>.<origin>.<operation>
```

| Segment | Codes |
|---|---|
| accommodation | `552001` holiday apartments/homes, `552002` group accommodation |
| indicator | `20101` arrivals, `20201` overnight stays, `20401` length of stay |
| origin | `tot` total, `ch` Switzerland, `ausl` foreign countries |
| operation | `sumw` weighted sum, `valuew` weighted value, `ci` confidence interval (+/-) |

The cube is sparse: arrivals and overnight stays come as `sumw` with a `ci`,
length of stay as `valuew` only. Labels for every code, in German, French,
Italian and English, are in `data-raw/metadata.yaml`.

## A note on vintages

OpenTSI archives use git history as the vintage archive. stats.swiss publishes
only the *current* vintage, so unlike `opentsi/ch.fso.besta` this archive has no
imported backlog: its history starts at the first automated commit and
accumulates revisions from there.

## Browse Time Series Data

You can use GitHub's ability to render csv to explore the datasets — see
[`data-raw/index.md`](data-raw/index.md).

## Basic Data Consumption via opentimeseries

```{r eval=FALSE}
remotes::install_github("opentsi/opentimeseries")
library(opentimeseries)

# first param `series` defaults to NULL and fetches all series
ts <- read_open_ts(
  remote_archive = "benjamin-arfa/ch.fso.pasta"
)

ts
```

```{r eval=FALSE}
library(opentimeseries)
library(tsbox)

# total overnight stays in holiday apartments
ts <- read_open_ts(
  series = "552001.20201.tot.sumw",
  remote_archive = "benjamin-arfa/ch.fso.pasta"
)

ts$date <- as.Date(ts$date)
ts_plot(ts)
```
